
A simple way to connect a model to OpenDA is by letting OpenDA access the input
and output files of the model. OpenDA cannot directly understand the input and
output files of an arbitrary model. Some code has to be written such that the
black box model implementation of OpenDA can read and write these files. In
this exercise you will learn how to connect an existing model to OpenDA
assuming that all the input and output files of the model can indeed be
accessed by OpenDA. The exercise focusses on the configuration of the black box
wrapper in OpenDA.\\

In the directory {\tt exercise\_4/stochModel/bin/} you will find a model written in python \\
{\tt reactive\_pollution\_model.py} and a compiled version of this code for
windows. There is also an input file ({\tt reactive\_pollution\_model.input})
and the output file you should get when you run the model. The model describes
the advection of two chemical substances. The first substance $c_1$ is emitted
as a pollutant by a number sources. However, in this case this substance reacts
with the oxygen in the air to form a more toxic substance $c_2$. The model
implements the following equations:

\begin{eqnarray}
    \frac{\partial c_1}{\partial t} + u\frac{\partial c_1}{\partial x} & = & -
    1/T c_1 \\
    \frac{\partial c_2}{\partial t} + u\frac{\partial c_2}{\partial x} & = &
    1/T c_1
\end{eqnarray}

\begin{itemize}
\item Run the model from the command line (passing input file as argument),
not using OpenDA.\\
The model
  generates the output files: {\tt reactive\_pollution\_model.output} and
  {\tt reactive\_pollution\_model\_output.m}. Use the m-file to make plots
  of the output in order to study the behavior of the model. In order to check
  the model (plotted) results you can look at the input file as well.
\end{itemize}

For this exercise, the Java-routines for reading and writing the input and
output files are already programmed. However, it is not necessary to program
this in Java. It is also possible to write your own conversion program (in any
programming language) to convert the input and output files of your model to a
format that OpenDA is able to handle.

When you are interested in the actual java code that parses the input and output files of this black box model. You can find it at \\
{\tt \$OPENDA/model\_reactive\_advection} in a source distribution of OpenDA.

A black box wrapper configuration usually consists of three xml files. For our
pollution model these files are:
\begin{enumerate}
   \item {\tt polluteWrapper.xml}: This file specifies the actions to performed
     when the model has to be run, and the files and related reader and writers
     that can be used to let OpenDA interact with the model.\\ This file
     consists of the parts:
     \begin{itemize}
        \item {\tt aliasDefinitions:} This is a list of strings that can be
          aliased in the other xml files. This helps to make the
          wrapperxml-file more generic. E.g. the alias definition
          \%outputFile\% can be used to refer to the output file of the model,
          without having to know the actual name of that output file.\\ Note
          the special alias definition \%instanceNumber\%. This will be
          replaced internally at runtime with the member number of each created
          model instance.
        \item {\tt run:} the specification of what commands need to be executed
          when the model is run.
        \item {\tt inputOutput:} the list of 'input/output objects', usually
          files, that are used to access the model, i.e. to adjust the model's
          input, and to retrieve the model's results. For each 'ioObject' one
          must specify:
        \begin{itemize}
           \item the java class that handles the reading from and/or writing to
             the file
           \item the identifier of the ioObject, so that the model
             configuration file can refer to it when specifying the model
             variables that can be accessed by OpenDA, the so called 'exchange
             items' (see below)
           \item optionally, the arguments that are needed to initialize the
             ioObject, i.e. to open the file.
        \end{itemize}
     \end{itemize}
   \item {\tt polluteModel.xml}: This is the main specification of the
     (deterministic) model. It contains the following elements:
     \begin{itemize}
        \item {\tt wrapperConfig}: A reference to the wrapper config file
          mentioned above.
        \item {\tt aliasValues}: The actual values to be used for the aliases
          defined in the wrapper config file. For instance the \%outputFile\%
          alias is set to the value "reactive\_pollution\_model.output".
        \item {\tt timeInfoExchangeItems}: The name of the model variables (the
          'exchange items') that can be accessed to modify the start and end
          time of the period to that the model should compute to propagate
          itself to the next analysis time.
        \item {\tt exchangeItems}: The model variables that are allowed to be
          accessed by OpenDA, for instance parameters, boundary conditions, and
          computed values at certain locations. Each variable exchange item
          consists of its id, the ioObject that contains the item, and the
          'element name', the name of the exchange item in the ioObject.
     \end{itemize}
   \item {\tt polluteStochModel.xml}: This is the specification of the
     stochastic model. It contains of two parts:
     \begin{itemize}
        \item {\tt modelConfig}: A reference to the deterministic model
          configuration file mentioned above {\tt polluteModel.xml}.
        \item {\tt vectorSpecification}: The specification of the vectors that
          will be accessed by the OpenDA algorithm. These vectors are grouped
          in two parts:
          \begin{itemize}
             \item The state that is manipulated by an OpenDA filtering
               algorithm, i.e. the state of the model combined with the noise
               model(s).
             \item The so called predictions, i.e. the values on observation
               locations as computed by the model.
          \end{itemize}
     \end{itemize}
\end{enumerate}

Start with a single OpenDA-run to understand where the model results appear
for this configuration:
\begin{itemize}
 \item Have a look at the files {\tt polluteWrapper.xml}, {\tt
   polluteModel.xml} and {\tt polluteStochModel.xml}, and recognize the various
   items mentioned above. Start the OpenDA GUI from the {\tt public/bin}
   directory and run the model by using the {\tt Simulation.oda} configuration.
   Note that the actual model results are available in the directory where the
   black box wrapper has let the model perform its computation: {\tt
     stochModel/output/work0}.
\end{itemize}

In this exercise, we will calibrate the value of the reaction-rate constant.
The algorithm used in this example is the Dud (which stands for Doesn't Use
Derivative).

\begin{itemize}
\item Have a look at the {\tt Dud.oda} and the configuration files it refers
  to. Run it from the OpenDA GUI and have a look at the results. What could you
  do to improve the results?

\item Figure out where to change the control parameters for the calibration
  procedure and play around with the settings to improve your results.

\end{itemize}

Calibration runs normally take longer than a few minutes. In that case, it
becomes convenient to be able to restart from a previous run.

\begin{itemize}
\item Adapt the configuration in such a way that you are able to restart the
  Dud.oda from the result of a previous run.
\end{itemize}
