{\bf Directory: {\tt exercise\_black\_box\_steady\_state\_filter}}\\

  In this section you will learn how to create and use a steady-state Kalman
  filter with OpenDA. The example continuues with the black-box reaction-pollution-model. Make sure you have completed the previous exercise before you start with this one.

  The steady-state Kalman filter is a special case of the Kalmann filter. If the measurement stations are fixed in time and measure with a fixed time-step and the model is linear and time-invariant then the Kalman gain matrix can converge over time to a fixed matrix. Sufficient conditions for this are the stability of the model or the controllability and observability of the model, If the Kalman gain converges, then it's possible to compute it, store it and the re-use the stored gain-matrix. This has a large impact on the computational demand of the algorithm. The steady-filter only uses only a bit more computer time than the model, which is much faster than the Ensemble Kalman filter. The steady-state filter is therefore very suitable for real-time applications, when it can be applied. In a strict sense the steady-state filter has a limited applicability, but if the model is not very non-linear or well constrained by the observations, then the Kalman gain matrix can still show very little variablity over time. In this case the steady-state filter can still be used. Clearly, this will not always work, but the computational advantage are so large that it is often worth considering.


\begin{itemize}
\item Look at the implementation of the model in {\tt
  WaveStochModelInstance.java}, in the directory {\tt
  simple\_wave\_model/java/src/org/openda/}. See how the state is defined and how
  the model is discretized. If you want you can compile the model using {\tt ant
  build} as we will explain in excercise 6. However to make it easy for you, you
  will find the compiled version of this model,{\tt simple\_wave\_model.jar} in
  the directory {\tt simple\_wave\_model/bin}.
\item The model represents a "user" model that is not part of the OpenDA distribution. Therefore you have to copy the
 model jar-file to the bin directory of your OpenDA installation. In this way OpenDA can find this model.
\item Run the model ({\tt waveSimulation.oda}) and visualize the model results
  ({\tt plot\_movie.m} or {\tt plot\_movie.py}). Do not forget to add the jar-file of the model to the
  {\tt CLASSPATH} variable, or to copy the jar-file into the bin directory of
  your OpenDA version
\item Adjust the input files in order to run the model with stochastic
  forcings.
\item Generate water level observations from this stochastic run. We need
  observations at (approximately) $x=\frac{1}{4} L$, $x=\frac{1}{2} L$ and
  $x=\frac{3}{4} L$. You can use the script {\tt generate\_obs.m} for this
  task. We want to have observations at $t=0.1, 0.2,...,10.0$, (initially)
  select a standard deviation of 0.05.
\item Run the Ensemble Kalman filter ({\tt waveEnkf.oda}). This run will
  generate and write gain matrices at specified times. Find where and how this
  is specified in the input.
\item Plot the columns of the gain matrices. (The script {\tt plot\_gains.m}
  or {\tt plot\_gains.py}
  plots the water level part of the gain matrices). What do these columns mean?
\item (Re)generate the gain matrices using different numbers of ensembles. When
  you compare the gain matrices, what do you notice. Note: The algorithm will
  generate an enormous amount of output when you run the EnKF with a very large
  number of ensembles (e.g. 500). You can suppress the output by commenting out
  (or remove) the {\tt resultWriter}-part of the oda-input file.
\item Use the generated steady state gain matrices for a steady state Kalman
  run ({\tt waveSteadystate.oda}). Compare the performance of:
  \begin{itemize}
  \item a (non-stochastic) run without filtering,
  \item an EnkF run with various numbers of ensembles (do not forget to
    reinstate the resultWriter if you have switched it off),
  \item the various steady state gains.
  \end{itemize}
 you can use the scripts {\tt plot\_obs\_sim.m}, {\tt plot\_obs\_ens.m}
 and\\ {\tt plot\_obs\_steady.m} and similar routines for python.

\item Generate (observations) gain matrices but now for only a single
  observation. Make sure that the observed values are exactly the same as in
  the 3 observation observer. Compare the columns of the 3-observation gain
  matrices to the single observation matrices. What is the main difference and
  why?
\end{itemize}
