

A simple way to connect a model to OpenDA is by letting OpenDA access the input
and output files of the model. OpenDA cannot directly understand the input and
output files of an arbitrary model. Some code has to be written such that the
black box model implementation of OpenDA can read and write these files. In
this exercise you will learn how to connect an existing model to OpenDA
assuming that all the input and output files of the model can indeed be
accessed by OpenDA. The exercise focusses on the configuration of the black box
wrapper in OpenDA.\\

The model describes
the advection of two chemical substances. The first substance $c_1$ is emitted
as a pollutant by a number sources. However, in this case this substance reacts
with the oxygen in the air to form a more toxic substance $c_2$. The model
implements the following equations:

\begin{eqnarray}
    \frac{\partial c_1}{\partial t} + u\frac{\partial c_1}{\partial x} & = & -
    1/T c_1 \\
    \frac{\partial c_2}{\partial t} + u\frac{\partial c_2}{\partial x} & = &
    1/T c_1
\end{eqnarray}


In the directory {\tt { \opgave}/original\_model/} you will find:
\begin{enumerate}
	\item model executable {\tt reactive\_pollution\_model.py} (linux and mac)
	      and  {\tt reactive\_pollution\_model.exe} (windows)
	\item {\tt config.yaml} the model configuration file
        \item {\tt forcings} the forcings of the model (injection of pollutant)
	\item {\tt input} the initial model state 
\end{enumerate}


\begin{itemize}
	\item Run the model, in the {\tt original\_model} directory from the
	      command line, not using OpenDA.
\end{itemize}
The model generates the output files {\tt c1\_locA},{\tt c1\_locB},
{\tt c1\_locC}, {\tt c1\_locA}, {\tt c2\_locB}, {\tt c2\_locC} with timeseries
of substance $c_1$ and $c_2$ at three predefinined locations in the model. The
folders {\tt maps} contains output files with the conentration of $c_1$ and $c_2$ on each grid point at specfied times.
The folder {\tt restart} contains files that allows the model to restart; continue the computations from the point where a restart file has been written.
\begin{itemize}
	\item Investigate the input and output files of the model 
	\item generate a movie by running the script {\tt plot\_movie\_orig.py} script from the {\tt { \opgave}} (!) directory. This allows you  to study the behaviour of the model. 
\end{itemize}

\subsection{Wrapper configuration files}

The input and outputfiles of this model are all easy to interpret ASCII files. 
Therefore we do not need model specific routines to couple this model to OpenDA.

When you couple an arbitrary model to OpenDA and you want to use the black box coupler there are two appoaches:
\begin{itemize}
	\item write a pre and post processing sript that translates the (relevant)
	      model files into a more generic format that is already supported
	      (e.g. ASCII or NetCDF).
	\item write your own adapter in java (dataobject) to read and write the
	      model input and output files.
\end{itemize}

A black box wrapper configuration usually consists of three xml files. For our
pollution model these files are:
\begin{enumerate}
   \item {\tt polluteWrapper.xml}: This file specifies how OpenDA can run the model, which input and output files are involved and dataobjects are used to interpret the model files.
     \\ This file
     consists of the parts:
     \begin{itemize}
        \item {\tt aliasDefinitions:} This is a list of strings that can be
          aliased in the other xml files. This helps to make the
          wrapperxml-file more generic. E.g. the alias definition
          \%outputFile\% can be used to refer to the output file of the model,
          without having to know the actual name of that output file.\\ Note
          the special alias definition \%instanceNumber\%. This will be
          replaced internally at runtime with the member number of each created
          model instance.
        \item {\tt run:} the specification of what commands need to be executed
          when the model is run.
        \item {\tt inputOutput:} the list of 'input/output objects', usually
          files, that are used to access the model, i.e. to adjust the model's
          input, and to retrieve the model's results. For each 'ioObject' one
          must specify:
        \begin{itemize}
           \item the java class that handles the reading from and/or writing to
             the file
           \item the identifier of the ioObject, so that the model
             configuration file can refer to it when specifying the model
             variables that can be accessed by OpenDA, the so called 'exchange
             items' (see below)
           \item optionally, the arguments that are needed to initialize the
             ioObject, i.e. to open the file.
        \end{itemize}
     \end{itemize}
   \item {\tt polluteModel.xml}: This is the main specification of the
     (deterministic) model. It contains the following elements:
     \begin{itemize}
        \item {\tt wrapperConfig}: A reference to the wrapper config file
          mentioned above.
        \item {\tt aliasValues}: The actual values to be used for the aliases
          defined in the wrapper config file. For instance the \%configFile\%
          alias is set to the value "config.yaml".
        \item {\tt timeInfoExchangeItems}: The name of the model variables (the
          'exchange items') that can be accessed to modify the start and end
          time of the period to that the model should compute to propagate
          itself to the next analysis time.
        \item {\tt exchangeItems}: The model variables that are allowed to be
          accessed by OpenDA, for instance parameters, boundary conditions, and
          computed values at certain locations. Each variable exchange item
          consists of its id, the ioObject that contains the item, and the
          'element name', the name of the exchange item in the ioObject.
     \end{itemize}
   \item {\tt polluteStochModel.xml}: This is the specification of the
     stochastic model. It contains of two parts:
     \begin{itemize}
        \item {\tt modelConfig}: A reference to the deterministic model
          configuration file mentioned above {\tt polluteModel.xml}.
        \item {\tt vectorSpecification}: The specification of the vectors that
          will be accessed by the OpenDA algorithm. These vectors are grouped
          in two parts:
          \begin{itemize}
             \item The state that is manipulated by an OpenDA filtering
               algorithm, i.e. the state of the model combined with the noise
               model(s).
             \item The so called predictions, i.e. the values on observation
               locations as computed by the model.
          \end{itemize}
     \end{itemize}
\end{enumerate}

Start with a single OpenDA-run to understand where the model results appear
for this configuration:
\begin{itemize}
 \item Have a look at the files {\tt polluteWrapper.xml}, {\tt
   polluteModel.xml} and {\tt polluteStochModel.xml}, and recognize the various
   items mentioned above. Start the OpenDA GUI from the {\tt public/bin}
   directory and run the model by using the {\tt Simulation.oda} configuration.
   Note that the actual model results are available in the directory where the
   black box wrapper has let the model perform its computation: {\tt
     work/work0}.
\end{itemize}
